---
title: "Dengue competition"
author: "Sara Marín"
date: "12/09/2019"
output: 
  html_document:
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
      smooth_scroll: true
---

<html>
  <head>
  <style type="text/css">
    * {
      text-align: justify;
    }
    
    h2 {
      color: MidnightBlue;    
      font-weight: bold;
    }
    
    h3 {
      color: MidnightBlue;    
    }
    
    h4 {
       color: SteelBlue;    
     
    }
    
  </style>
  </head>
  <body>
  
This report is intended to explain the process that I followed in the **dengue competition**. The goal is to predict the number of dengue cases each week (in two different locations) based on environmental variables describing changes in temperature, precipitation, vegetation, and more.

You can read more about this competition [Here](https://www.drivendata.org/competitions/44/dengai-predicting-disease-spread/)

```{r eval=TRUE, echo=FALSE, warning=FALSE, include=FALSE}

# load Libraries: p_load can install, load,  and update packages

library(dplyr)
library(ggplot2)
library(lubridate)
library(plotly)
library(gdata)
library(here)

# # setwd
#setwd(here::here())
#
# Let's read the script with the own functions
#source("../script/functions_sara.R")
#
# # reading datasets
# data_x <- read.csv("./data/dengue_features_train.csv", stringsAsFactors = F)
# data_y<- read.csv("./data/dengue_labels_train.csv", stringsAsFactors = F)
# data<- cbind(data_x, total_cases=data_y$total_cases)
#
# validation<- read.csv("./data/dengue_features_test.csv", stringsAsFactors = F)
#
# rm(data_x, data_y)
```


```{r sourceCode,echo=TRUE}
# use source() function to source the functions we want to execute

```

## 1. EXPLORATORY ANALYSIS
### 1. Every variable ###
### 2. Relationship between variables 

## 2. CLEANING 
## 3. TRAINING
  

