---
title: "Dengue competition"
author: "Sara Marín"
date: "12/09/2019"
output: 
  html_document:
    code_folding: hide
    toc: true
    toc_depth: 4
    toc_float: 
      collapsed: false
      smooth_scroll: true
---

<html>
  <head>
  <style type="text/css">
    * {
      text-align: justify;
    }
    
    h2 {
      color: MidnightBlue;    
      font-weight: bold;
    }
    
    h3 {
      color: MidnightBlue;    
    }
    
    h4 {
       color: SteelBlue;    
     
    }
    
  </style>
  </head>
  <body>
  
This report is intended to explain the process that I followed in the **dengue competition**. The goal is to predict the number of dengue cases each week (in two different locations) based on environmental variables describing changes in temperature, precipitation, vegetation, and more.

You can read more about this competition [Here](https://www.drivendata.org/competitions/44/dengai-predicting-disease-spread/)

```{r eval=TRUE, echo=FALSE, warning=FALSE, include=FALSE}

# load Libraries: p_load can install, load,  and update packages

library(dplyr)
library(ggplot2)
library(lubridate)
library(plotly)
library(gdata)
library(here)
library(corrplot)
library(weathermetrics)
library(reshape2)
library(tidyr)
library(zoo)
library(formattable)

# reading datasets
data_x <- read.csv("dengue_features_train.csv", stringsAsFactors = F)
data_y<- read.csv("dengue_labels_train.csv", stringsAsFactors = F)
data<- cbind(data_x, total_cases=data_y$total_cases)

validation<- read.csv("dengue_features_test.csv", stringsAsFactors = F)

rm(data_x, data_y)

# Rename variables 
data <- data %>% rename("precip_amt"= "precipitation_amt_mm",
                        "precip_mm_r"= "reanalysis_sat_precip_amt_mm",
                        "temp_dewpoint_r"="reanalysis_dew_point_temp_k",
                        "temp_air_mean_r" = "reanalysis_air_temp_k",
                        "humid_relative_r" = "reanalysis_relative_humidity_percent",
                        "humid_specific_r" = "reanalysis_specific_humidity_g_per_kg",
                        "precip_kgperm2_r" = "reanalysis_precip_amt_kg_per_m2", 
                        "temp_max_r" = "reanalysis_max_air_temp_k",
                        "temp_min_r" = "reanalysis_min_air_temp_k",
                        "temp_air_avg_r" = "reanalysis_avg_temp_k", 
                        "temp_dir_range_r" = "reanalysis_tdtr_k",
                        "temp_max_st"= "station_max_temp_c",
                        "temp_min_st"= "station_min_temp_c",
                        "temp_avg_st"= "station_avg_temp_c",
                        "precip_st" = "station_precip_mm",
                        "temp_dir_range_st" = "station_diur_temp_rng_c")

# Rename variables 
validation <- validation %>% rename("precip_amt"= "precipitation_amt_mm",
                                    "precip_mm_r"= "reanalysis_sat_precip_amt_mm",
                                    "temp_dewpoint_r"="reanalysis_dew_point_temp_k",
                                    "temp_air_mean_r" = "reanalysis_air_temp_k",
                                    "humid_relative_r" = "reanalysis_relative_humidity_percent",
                                    "humid_specific_r" = "reanalysis_specific_humidity_g_per_kg",
                                    "precip_kgperm2_r" = "reanalysis_precip_amt_kg_per_m2", 
                                    "temp_max_r" = "reanalysis_max_air_temp_k",
                                    "temp_min_r" = "reanalysis_min_air_temp_k",
                                    "temp_air_avg_r" = "reanalysis_avg_temp_k", 
                                    "temp_dir_range_r" = "reanalysis_tdtr_k",
                                    "temp_max_st"= "station_max_temp_c",
                                    "temp_min_st"= "station_min_temp_c",
                                    "temp_avg_st"= "station_avg_temp_c",
                                    "precip_st" = "station_precip_mm",
                                    "temp_dir_range_st" = "station_diur_temp_rng_c")

# Transform some variables to factor/numeric/datetime
data$week_start_date<- ymd(data$week_start_date)
validation$week_start_date<- ymd(validation$week_start_date)

data$city <- as.factor(data$city)
data$year <- year(data$week_start_date)
data$weekofyear<- as.factor(data$weekofyear)
data$month<- month(data$week_start_date)
validation$city <- as.factor(validation$city)
validation$year<- year(validation$week_start_date)
validation$weekofyear <- as.factor(validation$weekofyear)
validation$month<- month(validation$week_start_date)
validation$total_cases<-0

# Combine train and validation. Then split per city
data_full<-gdata::combine(data, validation)
data_iq<-data_full %>% filter(city=="iq")
data_sj<-data_full %>% filter(city=="sj")

# FUNCTIONS
# Plots geom density + variable in facet_wrap (ex. source in facet_wrap, for 
# comparing training and validation)
geom.density.function<- function(data, variables, fillby){
  
  # Melt with the selected variables
  data<- data[variables] %>%
    melt(fillby) %>%
    rename("id"=fillby)
  
  # geom_density
  ggplot(data, aes(x = value, fill = id)) + 
    geom_density(alpha = 0.2) + facet_wrap(~variable, scales = "free_x")
  
}

# Plotly with lines. Example: datetime and different variables 
# (val and train together)
plotly.line.function <- function(data, variables, x){
  
  # Melt with the selected variables
  data<- data[variables] %>%
    melt(x) %>%
    rename("id"=x)
  
  # plotly line
  plot_ly(data, x = ~id, y = ~value, 
          color = ~variable, type = 'scatter', mode = 'lines') 

}

# REMOVE COLLINEARITY
# By default, the threshold is going to be 0.85

remove.collinearity<- function(data, indep_var, threshold=0.85){
  
  # 1.Check if all the variables are numeric
  if(sum(sapply(data, is.numeric))!=ncol(data)){
    print("You have non-numeric variables!")
  }
  
  # 2. Create the correlation matrix
  cor_indep_var<- cor(within(data, rm("total_cases")))
  cor_indep_var[upper.tri(cor_indep_var, diag=TRUE)] <- NA
  cor_indep_var<-as.data.frame(reshape2::melt(cor_indep_var, na.rm=TRUE, 
                                              value.name="cor"))

  cor_dep_var<- as.data.frame(cor(data)) %>% select(indep_var)
  
  # 3. Show collinearity
  cor_indep_var_toremove<-cor_indep_var %>% filter(cor>=threshold)
  
  # 4. Remove collinearity
  
  # Let's create an empty vector
  variables_to_remove<-c()
  
  for (i in 1:nrow(cor_indep_var_toremove)){
    
    # Variables to check in the correlation matrix with the dep. var.  
    v1<- cor_indep_var_toremove[i,"Var1"] 
    v2<- cor_indep_var_toremove[i,"Var2"]
    
    corv1<-abs(cor_dep_var[which(rownames(cor_dep_var)==v1),"total_cases"])
    corv2<-abs(cor_dep_var[which(rownames(cor_dep_var)==v2),"total_cases"])
    
    print(paste(v1, " = ", corv1, " ", v2, " = ", corv2))
    
    if(corv1 >= corv2){
      print(paste("Let's remove ", v2))
      variables_to_remove<-append(variables_to_remove, v2)
    
    } 
    
    else{
      print(paste("Let's remove ", v1))
      variables_to_remove<-append(variables_to_remove, v1)
    }
    print("____________________________________")
      
  }
  
  data <- data %>% select(-variables_to_remove)
}
```
## A. FIRST CLEANING 
### A.1. MISSING VALUES ###

Let's explore where our missing values are located. For now, let's use the **na.locf** function for replacing them (although we will need to use a better way in the near future).    
```{r eval=TRUE, warning=FALSE}
# where and how many missing values 

missing_values <- data_sj %>%
  filter(year>2000) %>%
  select_if(is.numeric) %>%
  gather(key = "key", value = "val") %>%
  mutate(is.missing = is.na(val)) %>%
  group_by(key, is.missing) %>%
  summarise(num.missing = n()) %>%
  filter(is.missing==T) %>%
  select(-is.missing) %>%
  arrange(desc(num.missing)) 

ggplot(missing_values, aes(x=key, y=num.missing, fill=key)) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle=60, hjust=1)) 

data_sj<-na.locf(data_sj)
data_iq<-na.locf(data_iq)
rm(missing_values)
```

## B. EXPLORATORY ANALYSIS

After renaming and transforming some variables (into factor/numeric/datetime), we start splitting the dataset in the two different cities. 

### B.1. TEMPERATURE ###
#### Distributions ####

Let's compare the temperature distributions in **Iquitos**
```{r eval=TRUE, warning=FALSE}
#### 2.2. TEMPERATURE IQ & SJ _____________________________________________ ####

# Get the temperature variables
temp_var<-names(data[grep("temp", names(data))])
temp_var_k<-c("temp_air_mean_r","temp_air_avg_r","temp_dewpoint_r","temp_max_r",
            "temp_min_r")

# Change everything to celsius
data_iq[temp_var_k]<- kelvin.to.celsius(data_iq[temp_var_k], round = 2)
data_sj[temp_var_k]<- kelvin.to.celsius(data_sj[temp_var_k], round = 2)

rm(temp_var_k)

# Temperature variables train & validation 
geom.density.function(data=data_iq, variables=c(temp_var[1:10], "source"), 
                      fill="source")
```

Let's compare the temperature distributions in **San Juan**

```{r eval=TRUE, warning=FALSE}
geom.density.function(data=data_sj, variables=c(temp_var[1:10], "source"), 
                      fill="source")
```

#### Relations ####

Let's plot together all temperature variables in **Iquitos**

```{r eval=TRUE, warning=FALSE}
# Plot the progression of the temperature variables
plotly.line.function(data_iq,variables= c(temp_var[1:10], "total_cases", 
                                          "week_start_date"), x="week_start_date")
```

Let's plot together all temperature variables in **San Juan** (we have divided by 10 the dependent variable for visualization purposes only)
```{r eval=TRUE, warning=FALSE}
datatemp_sj <- data_sj
datatemp_sj$total_cases<-datatemp_sj$total_cases/10

plotly.line.function(datatemp_sj,variables= c(temp_var[1:10], "total_cases", 
                                          "week_start_date"), x="week_start_date")

rm(temp_var, datatemp_sj)
```

### B.2. HUMIDITY & PRECIP ###
#### Distributions ####

Let's compare the humidity and precipitation distributions in **Iquitos**
```{r eval=TRUE, warning=FALSE}
# Get the humidity variables
humid_precip_var<-names(data[grep("humid|precip", names(data))])

# humidity variables train & validation 
geom.density.function(data=data_iq, variables=c(humid_precip_var, "source"), 
                      fill="source")
```

Let's compare the  humidity and precipitation distributions in **San Juan**

```{r eval=TRUE, warning=FALSE}
geom.density.function(data=data_sj, variables=c(humid_precip_var, "source"), 
                      fill="source")
```

#### Relations ####
Let's plot together all humidity and precipitation variables in **Iquitos**
```{r eval=TRUE, warning=FALSE}
# Plot the progression of the humid & precip variables
plotly.line.function(data_iq,variables= c(humid_precip_var, "total_cases", 
                                          "week_start_date"), x="week_start_date")
```

Let's plot together all humidity and precipitation variables in **San Juan**
```{r eval=TRUE, warning=FALSE}
# Plot the progression of the humid & precip variables
plotly.line.function(data_sj,variables= c(humid_precip_var, "total_cases", 
                                          "week_start_date"), x="week_start_date")
rm(temp_humid_precip)
```

### B.3. VEGETATION ###

#### Distributions ####
Let's compare the vegetation distributions in **Iquitos**
```{r eval=TRUE, warning=FALSE}
# Get the vegetation variables
var_veg<-names(data[grep("ndvi", names(data))])

# humidity variables train & validation 
geom.density.function(data=data_iq, variables=c(var_veg, "source"), 
                      fill="source")
```

Let's compare the vegetation distributions in **San Juan**
```{r eval=TRUE, warning=FALSE}
geom.density.function(data=data_sj, variables=c(var_veg, "source"), 
                      fill="source")
```

#### Relations ####
Let's plot together all vegetation variables in **Iquitos** (we have multiplied by 100 the vegetation variables for visualization purposes only)
```{r eval=TRUE, warning=FALSE}
# Plot the progression of the veg variables
dataveg_iq <- data_iq
dataveg_iq[var_veg]<-dataveg_iq[var_veg]*100
plotly.line.function(dataveg_iq,variables= c(var_veg, "total_cases", 
                                          "week_start_date"), x="week_start_date")
```

Let's plot together all vegetation variables in **San Juan** (we have multiplied by 1000 the vegetation variables for visualization purposes only)
```{r eval=TRUE, warning=FALSE}
dataveg_sj <- data_sj
dataveg_sj[var_veg]<-dataveg_sj[var_veg]*1000
plotly.line.function(dataveg_sj,variables= c(var_veg, "total_cases", 
                                          "week_start_date"), x="week_start_date")

rm(var_veg,dataveg_iq, dataveg_sj)
```

## C. REDUCING DIMENSIONALITY

### C.1. REMOVE COLLINEARITY ###
Let's remove those redundant predictors that are correlated with other predictors. In the table you can check those removed variables (using a threshold = 0.80)
```{r eval=TRUE, warning=FALSE, include=FALSE}
# Original column names
origvar_iq<-names(data_iq)
origvar_sj<-names(data_sj)

# Save categorical variables and create a df withouth them
cat_var<-c("week_start_date", "city", "year", "weekofyear", "source")
data_cor_matrix_iq<- data_iq %>% select(-cat_var) 
data_cor_matrix_sj<- data_sj %>% select(-cat_var) 

# Remove collinearity from data and validation
data_iq<-cbind(remove.collinearity(data=data_cor_matrix_iq,         # 25 <- 20
                                   indep_var="total_cases",
                                   threshold=0.80),data_iq[cat_var])

data_sj<-cbind(remove.collinearity(data=data_cor_matrix_sj,         # 25 <- 15
                                   indep_var="total_cases",
                                   threshold=0.80),data_sj[cat_var])


rm(cat_var, data_cor_matrix_iq,data_cor_matrix_sj)

```


```{r eval=TRUE, warning=FALSE}
final_variables<-data.frame(Iquitos = names(data_full), San_Juan = names(data_full))
rownames(final_variables)<- names(data_full)

final_variables$Iquitos<-ifelse(final_variables$Iquitos %in% names(data_iq), 
                                "Yes", "Removed")
final_variables$San_Juan<-ifelse(final_variables$San_Juan %in% names(data_sj), 
                                "Yes", "Removed")

# Replacing empty cells with icons
formattable(final_variables, list(
  Iquitos = formatter("span", style = x ~ ifelse(x == "Yes", 
    style(color = "green", font.weight = "bold"), style(color="red",font.weight = "bold"))),
  San_Juan = formatter("span", style = x ~ ifelse(x == "Yes", 
    style(color = "green", font.weight = "bold"), style(color="red",font.weight = "bold")))
  
))

```
  